NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-01-01 Manuscript/manuscriptPS.tex
Organize and update the manuscript

@@

\section{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}

@@

Absolute values of experimental order parameters for different lipid headgroups are
collected from the literature in Fig. \ref{HGorderParameters}. Since order parameter
signs are known only for PC, only absolute values are shown.
Main conclusions regarding the structure of different common lipid headgroups in the literature are \\
1) glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, \\
2) glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}, \\
3) headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. \\
Careful discussion and analysis of structural details of PE, PG or PS headgroups is not available, 
in contrast to PC lipids (see \cite{botan15} and references therein).
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparameters.eps}
  \caption{\label{HGorderParameters}
    Absolute values of order parameters for headgroup and glycerol backbone with different headgroups
    from experiments. POPC values are from \cite{ferreira13}, DOPS from \cite{browning80} contains 0.1M of NaCl,
    POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

@@

\subsection{Headgroup and glycerol backbone structure in simulations and experiments of PS lipid bilayers}
The headgroup order parameters of DOPS and POPS bilayers 
from different experiments and simulations are shown in
Figs. \ref{HGorderParametersPS}.
None of the tested models gives satisfactory agreement with
experiments for order parameters in headgroup $\alpha$ and $\beta$
carbons.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
  \todo{Check and report all the counterions.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
\end{figure}

Glycerol backbone order parameters seems similar in all models,
except in Slipids. 
Slipid model do not correctly capture the glycerol backbone
structure, as already observed previously for PC lipids \cite{botan15}.
The glycerol backbone structures between PC and
PS lipids simulated with Slipids are compared with the structures
simulated with CHARMM36 in Fig. \ref{glycerol_buslaev}.
The differences in sampled conformation leading to the order parameter
differences are clearly visible in the figure.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
  \caption{\label{glycerol_buslaev}
    Snapshots overlayed from different simulations for glycerol backbone region
    by Pavel Buslaev.
  }
\end{figure}

\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be included in the discussion.}

\subsection{Headgroup structure in PS and PC mixtures}
The headgroup order parameters of POPC 
mixed with PS lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
%The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
%order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
%in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
%headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
%smaller than in experiments.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    Headgroup order parameters from PC:PS mixtures from 
    different simulation models and experiments.
    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
    signs are determined as discussed in \cite{botan15,ollila16}).
    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
    Counterions in expriments are sodium, while potassium is used in simulations.
    Using sodium in simulations do not have a significant effect.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{We need results also from other than CHARMM36 force field.}
\end{figure*}

\todo{More data to be collected before discussion.}

%\subsection{Effect of PC on PS headgroup}
The headgroup order parameters of PS mixed with varying amounts of PC
from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
%The changes in PG headgroup order parameters are minor in simulations, which
%is in line with the only available experiment for the $\beta$-carbon.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
%  \caption{\label{HGorderparametersPSPGvsPC}
%    PS order parameters from mixtures with PC
%    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
%    Signs for PS are measures as described in SI.
%  }
%  \todo{Some simulations contain potassium as counterions, while some sodium.
%    All experiments here contain some amount of sodium salt. The best
%    ion concentrations for comparison should be figured out.} \\
%\end{figure}

>>>>

Glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90}.

@@

\subsection{Headgroup and glycerol backbone order parameters measured from POPS lipid bilayer}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are compared to the literature values of
DOPS \cite{browning80} and POPC \cite{ferreira13} in Fig. \ref{HGorderParameters}. 
Our results for POPS are in good agreement with the previously reported
values for DOPS measured with $^2$H NMR. Significant differences are
observed between PC and PS lipids, especially at the headgroup region.  
Previous discussions in the literature have concluded that  
glycerol backbone structure is largely similar in
PC, PE, PG and PS lipids \cite{gally81}. The headgroup region was
found to be similar in PC, PE and PG lipid (assuming that the signs of PE and PG
order parameters are the same as in PC), while the PS headgroup was suggested
to be more rigid \cite{wohlgemuth80,buldt81}. 
The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPS.eps}
  \caption{\label{HGorderParameters}
    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
    as measured in this work and signs for PC as measured in Refs \cite{??,ferreira16}.
    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
\end{figure}

@@

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$ 
    Experimental data from \cite{browning80} contains 0.1M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown for the computational values are not error bars, but demonstrate that for these systems we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
  \todo{Check and report all the counterions. There may be some simulation data 
    sets with both Na+ and K+ ions?}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
  \todo{CHARMM36-UA point for alpha in wrong place?}
\end{figure}

The headgroup order parameters of DOPS and POPS bilayers 
from different simulation models are compared with the experimental
data in Fig. \ref{HGorderParametersPS}. In line with the previous
study for PC lipids \cite{botan15}, the glycerol backbone order 
parameters in CHARMM36 and Berger force fields roughly agree with
the experimental data, but significant discrepancies are observed for
Slipids and CKP models. The glycerol backbone structures from CHARMM36 and
Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
missing structures in Slipids model, which probably lead to the incorrect
order parameters. For the headgroup $\alpha$ and $\beta$ carbon order parameters of PS lipids,
the tested models perform less well than for PC headgroup in previous
study \cite{botan15}. None of the tested models give a satisfactory 
agreement with experiments for the headgroup order paramaters of 
PS lipids. Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
the CHARMM36 is the best performing model for both lipids.
However, the total deviation from the experiments for the PS lipids is significantly
higher for PS lipids (8) than for PC (3) \cite{botan15}. Therefore the 
interpretation of structural differences between PC and PS lipids 
from current MD simulation models is very challenging.

\todo{Dihedral angle distribtions in Fig. \ref{dihedrals} should be analyzed from different models 
  and included in the discussion. Maybe also figures similar to \ref{glycerol_buslaev}.}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/glycerol_buslaev.png}
  \caption{\label{glycerol_buslaev}
    Snapshots overlayed from different simulations for glycerol backbone region
    by Pavel Buslaev.
  }
\end{figure}

\subsection{Headgroup structure in PS and PC mixtures}
Lipid mixtures containing PC and PS lipids are present in biological
systems. Also the Ca2+ binding affinity to bilayers containing
PS lipids is measured from mixtures with PC, because pure PS
bilayers exhibit instant transition to the ordered phase with the
addition of the ions. Therefore, the characterization of mutual interactions
between PC and PS lipid headgroups is necessary in order to
understand PS lipids in biological environments.

Fig. \ref{HGorderparametersPCvsPS} shows the headgroup order parameters
of POPC and POPS lipids in mixtures with various mole fractions from 
simulations and experiments \cite{scherer87,roux90}.
The headgroup order parameters of POPC increase with the increasing
amount of negatively charged POPS, as expected from the 
the electrometer concept \cite{seelig87}. Notably, the concept
seems to be generally valid, at least qualitatively, also for 
other lipid mixed with PC lipids; the PC lipid headgroup order
parameters increase when mixed with negatively charged 
lipids (PS, PI, CL, PA and PG) and remains almost unchaged when 
mixed with neutral lipids (PE and SM) \cite{scherer87} or cholesterol \cite{ferreira13}.
This is, however, not reproduced in the simulation data 
in Fig. \ref{HGorderparametersPCvsPS}. The order parameter for
$\beta$-carbon of POPC decreases and almost negligible increase is 
seen in $\alpha$ order parameter with the addition of PS lipid in CHARMM36 simulation.
This may indicate incorrect interactions between PC and PS headgroups
in simulations. On the other hand, the previously reported overestimated
ion binding affinity may also play a role, as the amount of counterions
is increasing with PS.
\todo{Data from other force fields to be collected before discussion can be finalized.}

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Headgroup order parameters from PC:PS mixtures from 
    different simulation models and experiments.
    Left panel shows the PC headgroup order parameters (experimental results from Ref. \citenum{scherer87},
    signs are determined as discussed in \cite{botan15,ollila16}).
    Rigth panel shows PS headgroup order parameters (experimental result for pure POPS measured in this
    work at 298K, experimental result for mixture from Ref. \citenum{roux90} at 298K).
    Counterions in expriments are sodium, while potassium is used in simulations.
    Using sodium in simulations do not have a significant effect.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{We need results also from other than CHARMM36 force field.}
\end{figure*}

Also the headgroup order parameters of POPS mixed with varying amounts of POPC
from simulations and experiments \cite{borle85,roux90} are shown in Fig. \ref{HGorderparametersPCvsPS}.
The $\beta$-carbon order parameter of POPS slightly decreases and positive 
order parameter of $\alpha$-carbon slightly increases with increasing
amount of PS lipids. This may indicate increasing order of the headgroup.
It should be, however, noted that the experimental data for pure POPS and
POPC/POPS mixture come from different experimental sets, 13C NMR in this work
and 2H NMR from Ref. \citenum{roux90}, respectively. Therefore the accuracy
of the order parameter change is not as high as typically in the measurements
of order parameter changes, see discussion about qualitative and quantitative 
accuracy in Ref. \citenum{ollila16}. The changes of PS headgroup order parameters
are not reproduced by the tested simulation models. The $\beta$-carbon order parameter
increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
to the experimental data. The smaller $\alpha$-carbon order parameter increase in both
simulation models with increasing amount of PS, while it is almost unchanged in experiments.
The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
with increasing amount of PS, both model exhibiting a poor agreement with experiments. 
Signifincant improvement in the MD simulation models are needed to interpret the
PS headgroup structures and mutual interactions with PC lipids.
